---
title: "ImpulseDE"
author: "Jil Sander"
date: "13 November 2015"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{ImpulseDE}
  %\VignetteEngine{knitr::rmarkdown}
---

# Table of Contents
1. [Introduction](#introduction)  
1.1 [Requirements for input data](#requirements-for-input-data)  
1.2 [Calling ImpulseDE](#calling-impulsede)  
1.3 [Output of ImpulseDE](#output-of-impulsede)
2. [Differential expression analysis for single time courses](
#differential-expression-analysis-for-single-time-courses)  
2.1 [Running the analysis 1TK](#running-the-analysis-1tk)  
2.2 [Plotting genes of interest 1TK](#plotting-genes-of-interest-1tk)  
2.3 [Imputing values for missing time points 1TK](
#imputing-values-for-missing-time-points-1tk)
3. [Differential expression analysis between two time courses](
#differential-expression-analysis-between-two-time-courses)  
3.1 [Running the analysis 2TK](#running-the-analysis-2tk)  
3.2 [Plotting genes of interest 2TK](#plotting-genes-of-interest-2tk)  
3.3 [Imputing values for missing time points 2TK](
#imputing-values-for-missing-time-points-2tk)
4. [References](#references)

# Introduction

*ImpulseDE* detects differentially expressed (DE) genes in high-throughput
time course experiments. It accepts two different kinds of inputs: 
whether a single time course dataset (1TK) or a dataset containing a case as
well as a control time course for each time point measured (2TK). 
For the first scenario, it identifies genes being differentially
expressed across time points. For the second scenario, *ImpulseDE* reports 
genes being differentially expressed between both conditions.
*ImpulseDE* follows a five-step workflow: 

------------------------------------------------------------------------------
Step               Explanation  
-----------------  ----------------------------------------------------------
Clustering         The genes are clustered into a limited number of groups  
                   using k-means. In default modus, *ImpulseDE* generates PDF   
                   documents, which contain plots for each cluster.   
                   Additionally, a text-file is produced containing each gene  
                   together with its corresponding cluster number.  
                  
Fit to clusters    *ImpulseDE* is based on the impulse model proposed by 
                   Chechik and Koller, which  reflects a two-step behavior of 
                   genes within a cell responding to environmental changes 
                   [@chechikkoller2009]. This model is fitted to the mean
                   expression profiles of the clusters.  

Fit to genes       The best parameter sets obtained from the clusters are then
                   used to fit an impulse model to each gene. 

Fit to random data The impulse model is fitted to a randomized dataset
                   (bootstrap), which is essential to detect
                   differentially expressed genes [@storey2005].  

Detection of       Detection of differentially expressed genes utilizing the
differentially     fits to the real and randomized data sets. FDR-correction
expressed genes    is performed to obtain adjusted p-values [@bh1995].
------------------------------------------------------------------------------

After each of those steps, a workspace containing the corresponding result
object is saved into the working directory.

## Requirements for input data
*ImpulseDE* requires an expression table (only numbers) as well as an annotation
table (characters allowed). The requirements for the two tables are the 
following:  

--------------------------------------------------------------------------------
Feature            Explanation  
-----------------  -------------------------------------------------------------
Expression table   Genes have to be in rows and samples in columns. Both rows
                   and columns should have unique identifiers.  
                   
Annotation table   Must have two columns, one carrying the timestamps as numeric
                   numbers and the other one carrying the condition information.
                   In the case of two time courses, two conditions are required.
                   More than two conditions are allowed to be specified, but
                   then two conditions (one case and one control conditions)
                   have to be specified by the user for each run separately. The
                   samples (row names) do not have to have the same order as in
                   the expression table (column names), but the sample
                   identifiers must be identical. Additional columns are allowed
                   but will be ignored later on.  
                   
Missing values     Are **not supported**. Genes having missing values for at
                   least one sample will be excluded from the analysis.  
                   
Time points        Since the parametric model contains six parameters, the
                   dataset should contain at least **six** time points.  
                   
Normalization      Gene expression data should be **properly normalized**
and filtering      including **log2-transformation** and **filtered** to avoid
                   spending time on fitting the model to non-informative
                   genes (e.g. not expressed or not variable genes). No impulse
                   model will be fitted to genes having a coefficient of
                   variation less than 0.025; instead, the mean across all
                   samples is returned as the "fit".
--------------------------------------------------------------------------------


## Calling ImpulseDE
The first four input parameters have to be specified by the user. Those are the
names of the two input tables (*expression_table* and *annotation_table*) as
well as the two column names carrying the time (*colname_time*) and condition
(*colname_condition*) information within the annotation table. Additional
parameters can be set to specify the time course scenario as well as fitting
and parallelization parameters.  
In the default modus, *ImpulseDE* expects a single time course scenario
without any control data (*control_timecourse = FALSE* and *control_name = 
NULL*). In the case of two time courses, the *control_timecourse* parameter has
to be set to *TRUE* and for *control_name* the name of the control within
*colname_condition* has to be specified. If more than two conditions are
present within the annotation table, *case_name* has to be set in
addition to run *ImpulseDE* for the desired case condition.  
Regarding the fitting, as default *ImpulseDE* will run 100 iterations 
(*n_iter = 100*) to optimize the model parameters, generate 50.000 random data
points (*n_randoms = 50.000*) to estimate bootstrapped p-values for DE analysis,
and determines DE genes using an FDR-adjusted p-value cutoff (q-value) of 1%
(*Q_value = 0.01*). Furthermore, in default modus it will split the run into 4
processes (*n_process = 4*). If parallelization is not possible on the device or
is not admired, *n_process* should be set to *1*. 

## Output of ImpulseDE

*ImpulseDE* returns a list consisting of two sublists: *impulse_fit_results* and
*DE_results*. The former contains the fitted impulse model parameters, sum of
squared fitting errors as well as the calculated impulse values for all time
points. The latter contains the names of the genes being called as
differentially expressed according to a specified cutoff together with the
adjusted p-values. Additionally, *ImpulseDE* saves the results from clustering
and DE analysis as text files and resulting plots within PDF files.

# Differential expression analysis for single time courses
In the case of a single time course experiment, *ImpulseDE* will detect
differentially expressed genes over time. A fitting dataset is provided within
the R package *longitudinal*, where T cells were stimulated with PMA and
ionomicin and harvested at 10 different time points [@rangel2004]. The dataset
contains 10 measurements per time point for 58 genes:

```{r}
# Install package longitudinal and load it
library(longitudinal)
# attach T cell data. 
data(tcell)
# check dimension of data matrix of interest
dim(tcell.10)
```

In order to be able to apply *ImpulseDE* on this dataset, *tcell.10* has to be
transposed using *t()* during the call since genes need to be in rows and
samples in columns. Additionally, it is necessary to create a proper
annotation table:

```{r}
# generate annotation table with columns "Time" and "Condition"
annot <- as.data.frame(cbind("Time" =
   sort(rep(get.time.repeats(tcell.10)$time,10)),
   "Condition" = "activated"), stringsAsFactors = FALSE)
# Time columns must be numeric
annot$Time <- as.numeric(annot$Time)
# rownames of annotation table must appear in data table
rownames(annot) = rownames(tcell.10)
head(annot)
```

It is important that the *Time* column contains numeric values and that the
*Condition* column is not a factor. Since the dataset contains only a sinlge
time course and therefore only one condition, the *Condition* column contains
only one unique value, *activated*.

## Running the analysis 1TK
*ImpulseDE* provides a single function, *impulse_DE*, which runs all the
analysis steps automatically and prints the current status on the screen. To run
*ImpulseDE* with all default options, only four variables need to be set for the
single time course scenario: *expression_table*, *annotation_table*, 
*colname_time* and *colname_condition*. However, for demonstration purposes the
number of randomizations as well as the number of used processors will be
reduced. For real datasets, it not recommended to reduce *n_randoms*.

```{r}
# load package
library(ImpulseDE)
# start analysis
impulse_results <- impulse_DE(t(tcell.10), annot, "Time", "Condition",
   n_randoms = 10, n_process = 1)
```

```{r echo = FALSE}
# rename plot files so that they are not overwritten during the two time course
# scenario
#file.rename("clusters_case_genes.pdf", "clusters_case_genes_1tk.pdf")
#file.rename("impulse_fit_genes_DE.pdf", "impulse_fit_genes_DE_1tk.pdf")
```

All indentified differentially expressed genes were plotted into
*impulse_fit_genes_DE.pdf* and the top 9 are shown within *Figure 1*.

<div style="text-align:center" markdown="1">
![](impulse_fit_genes_DE_1tk.pdf)
*Figure 1: Plots of the top 9 differentially expressed genes over time for the
T cell single time course dataset*
</div>

For example, *JUND* and *CD69* show very typical impulse-like expression
patterns, which cleary change significantly over time.

## Plotting genes of interest 1TK

*ImpulseDE* plots up to the top 36 DE genes automatically into the PDF-file
*impulse_fit_genes_DE.pdf*. If it is of interest to plot a custom list of genes,
this can be done using the function *plot_impulse*. As an example, four genes
will be plotted, which were excluded from the fitting process due to a very low
variation across the samples. For this the fitting results are needed, which
can be taken from the generated result object *impulse_results* or from the
saved workspace. *Figure 2* shows the resulting plots.

```{r}
load("impulse_fit_genes.RData")
genes = c("MYD88","CLU","CASP7","RBL2")
plot_impulse(gene_IDs = genes, data_table = t(tcell.10), data_annotation = annot,
    imp_fit_genes = impulse_fit_genes, file_name_part = "four_NV_genes")
```

<div style="text-align:center" markdown="1">
![](impulse_fit_genes_four_NV_genes.pdf)
*Figure 2: Plots of four almost non-variable genes from T cell single 
time course dataset*
</div>

## Imputing values for missing time points 1TK
To impute values for an uncovered time point for a specific gene, the following
command can be used:

```{r}
# impute expression value for time point 60 for gene "JUND"
(imp_results <- calc_impulse(impulse_fit_genes$impulse_parameters_case["JUND",
    1:6], 60))
```

# Differential expression analysis between two time courses

In the case of a two time course experiment, *ImpulseDE* will detect
differentially expressed genes between both conditions. In order to generate
two time courses out of the T cell data set, the replicates will be splitted
and to the second half some random numbers are added:


```{r}
# split dataset into two halfs
case_data <-  t(tcell.10)[,seq(1,ncol(t(tcell.10)),2)]
control_data <-  t(tcell.10)[,seq(2,ncol(t(tcell.10)),2)]
# add some random values to "control_data" to make data different
control_data <- control_data + t(apply(control_data,1,function(x) 
    runif(length(x),0,0.5)*sample(c(-1,1),length(x), replace = TRUE)
    + sample(c(seq(-2,2,0.5)),1)))
tcell_2tk <- cbind(case_data, control_data)

```

At last, a proper annotation table has to be generated:

```{r}
annot_2tk <- annot[colnames(tcell_2tk),]
annot_2tk[51:100,"Condition"] = "control"
head(annot_2tk)
tail(annot_2tk)

```

## Running the analysis 2TK
In contrast to the single time course scenario, six variables need to be set:
*expression_table*, *annotation_table*, *colname_time*, *colname_condition*, 
*control_timecourse* and *control_name*. Here again, for demonstration purposes
the number of randomizations as well as the number of used processors will be
reduced. For real datasets, it not recommended to reduce *n_randoms*.


```{r}
# load package
library(ImpulseDE)
# start analysis
impulse_results <- impulse_DE(tcell_2tk, annot_2tk, "Time", "Condition",
    TRUE, "control", n_randoms = 10, n_process = 1)
```

*Figure 3* shows the top 9 differentially expressed genes between the generated
case and control datasets.

<div style="text-align:center" markdown="1">
![](impulse_fit_genes_DE.pdf)
*Figure 3: Plots of the top 9 differentially expressed genes over time for the
T cell single time course dataset*
</div>

## Plotting genes of interest 2TK
As an example, the same four genes will be plotted as for the single time course
scenario. They are shown in *Figure 4*.

```{r}
load("impulse_fit_genes.RData")
genes = c("MYD88","CLU","CASP7","RBL2")
plot_impulse(gene_IDs = genes, data_table = tcell_2tk, data_annotation = 
    annot_2tk, imp_fit_genes = impulse_fit_genes, control_timecourse = TRUE,
    control_name = "control", file_name_part = "four_NV_genes_2tk")
```

<div style="text-align:center" markdown="1">
![](impulse_fit_genes_four_NV_genes_2tk.pdf)
*Figure 4: Plots of four almost non-variable genes from a two time course
dataset generated from the T cell data*
</div>

## Imputing values for missing time points 2TK
To impute values for an uncovered time point for a specific gene, *calc_impulse*
has to be applied to both datasets separately:

```{r}
# impute expression value for time point 60 for gene "JUND"
# case data
(imp_results <- calc_impulse(impulse_fit_genes$impulse_parameters_case["JUND",
    1:6], 60))
# control data
(imp_results <- calc_impulse(impulse_fit_genes$impulse_parameters_control[
    "JUND",1:6], 60))
```

# References

---
references:
- id: chechikkoller2009
  title: Timing of Gene Expression Responses to Environmental Changes.
  author:
  - family: Chechik
    given: Gal
  - family: Koller
    given: Daphne
  container-title: Journal of Computational Biology
  volume: 16
  URL: http://online.liebertpub.com/doi/abs/10.1089/cmb.2008.13TT
  DOI: 10.1089/cmb.2008.13TT
  issue: 2
  page: 279-290
  type: article-journal
  issued:
    year: 2009
    month: 2
- id: storey2005
  title: Significance analysis of time course microarray experiments.
  author:
  - family: Storey
    given: John D.
  - family: Xiao
    given: Wenzhong
  - family: Leek
    given: Jeffrey T.
  - family: Tompkins
    given: Ronald G.
  - family: Davis
    given: Ronald W.
  container-title: Proceedings of the National Academy of Sciences
  volume: 102
  URL: http://www.pnas.org/content/102/36/12837.full
  DOI: 10.1073/pnas.0504609102
  issue: 36
  page: 12837-12842
  type: article-journal
  issued:
    year: 2005
    month: 2
- id: bh1995
  title: >
   Controlling the False Discovery Rate: A Practical and Powerful
   Approach to Multiple Testing.
  author:
  - family: Benjamini
    given: Yoav
  - family: Hochberg
    given: Yosef
  container-title: > 
   Journal of the Royal Statistical Society Series B
   (Methodological)
  volume: 57
  URL: http://www.jstor.org/stable/2346101
  DOI: 10.2307/2346101
  issue: 1
  page: 289-300
  type: article-journal
  issued:
    year: 1995
    month: 3
- id: rangel2004
  title: >
   Modeling T-cell activation using gene expression profiling and
   state-space models.
  author:
  - family: Rangel
    given: Claudia
  - family: Angus
    given: John
  - family: Ghahramani
    given: Zoubin
  - family: Lioumi
    given: Maria
  - family: Sotheran
    given: Elizabeth
  - family: Gaiba
    given: Alessia   
  - family: Wild
    given: David L.   
  - family: Falciani
    given: Francesco
  container-title: Bioinformatics
  volume: 20
  URL: http://bioinformatics.oxfordjournals.org/content/20/9/1361.long
  DOI: 10.1093/bioinformatics/bth093
  issue: 9
  page: 1361-1372
  type: article-journal
  issued:
    year: 2004
    month: 2
---
